---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rmcp

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/rmcp)](https://CRAN.R-project.org/package=rmcp)
<!-- badges: end -->

The goal of rmcp is to enable LLM-enabled tools like Claude Code to learn about the R packages you have installed using the [Model Context Protocol](https://modelcontextprotocol.io/) (MCP). Tools configured with rmcp can peruse package documentation and learn to use R packages even if they weren't included in their training data.

rmcp is written primarily in TypeScript but is distributed as an R package for ease of distribution/install for R users. Installing the package ensures you have the needed dependencies and provides a shortcut, `mcp_config()`, to help you configure the tool with other applications.

> IMPORTANT:
> This is an early proof of concept. Use at your own risk!

## Installation

You can install the development version of rmcp like so:

```r
pak::pak("simonpcouch/rmcp")
```

rmcp can be hooked up to any application that supports MCP. Use `mcp_config()` to generate the config for your machine for common applications. For example, **Claude Code**:

```{r}
library(rmcp)

mcp_config("Claude Code")
```

Or, **Claude Desktop**:

```{r}
mcp_config("Claude Desktop")
```

## Example

```{r}
#| echo: false
#| fig-alt: "A screencast of a chat with Claude. I ask 'Do I have any R packages installed with the word flights in the name?' and, after searching through the documentation of my currently installed R packages, Claude mentions two that I have installed."
knitr::include_graphics("https://github.com/user-attachments/assets/8cefbc28-f046-4dfa-af63-b8eb85bb16b0")
```
