---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# acquaint

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/acquaint)](https://CRAN.R-project.org/package=acquaint)
[![R-CMD-check](https://github.com/simonpcouch/acquaint/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/simonpcouch/acquaint/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of acquaint is to enable LLM-enabled tools like Claude Code to learn about the R packages you have installed using the [Model Context Protocol](https://modelcontextprotocol.io/) (MCP). Tools configured with acquaint can peruse package documentation and learn to use R packages even if they weren't included in their training data.

acquaint is written primarily in TypeScript but is distributed as an R package for ease of distribution/install for R users. Installing the package ensures you have the needed dependencies and provides a shortcut, `mcp_config()`, to help you configure the tool with other applications.

> IMPORTANT:
> This is an early proof of concept. Use at your own risk!

## Installation

You can install the development version of acquaint like so:

```r
pak::pak("simonpcouch/acquaint")
```

acquaint can be hooked up to any application that supports MCP. Use `mcp_config()` to generate the config for your machine for common applications. For example, **Claude Code**:

```{r}
library(acquaint)

mcp_config("Claude Code")
```

Or, **Claude Desktop**:

```{r}
mcp_config("Claude Desktop")
```

## Example

In Claude Desktop, I'll write the following:

> Using the R packages I have installed, write code to download data on flights in/out of Chicago airports in 2024.

In a typical chat interface, I'd be wary of two failure points here:

1) The model doesn't know which packages I have installed.
2) If the model correctly guesses which packages I have installed, there may not be enough information about how to _use_ the packages baked into its weights to write correct code.

```{r}
#| echo: false
#| fig-alt: "A screencast of a chat with Claude. I ask 'Using the R packages I have installed, write code to download data on flights in/out of Chicago airports in 2024.' and, after searching through the documentation of my currently installed R packages, Claude writes R code to do so."
knitr::include_graphics("https://github.com/user-attachments/assets/821ea3d6-4e30-46d6-ab9b-301276af2c35")
```

Through first searching through my installed packages, Claude can locate the anyflights package, which seems like a reasonable solution. The model then discovers the  package's `anyflights()` function and reads its documentation, and can pattern-match from there to write the correct code.
